DateLife: leveraging databases and analytical tools to reveal the dated Tree of Life

### 2023-05-11 manuscript/manuscript.Rmd
add EJM edits

authornote: |
  School of Natural Sciences, University of California, Merced, 258 Science and Engineering Building 1, Merced, CA 95340, USA.

@@

Besides binomial species names, DateLife accepts scientific names from any inclusive taxonomic group (e.g., genus, family, tribe), as well as subspecific taxonomic variants (e.g., subspecies, variants, strains). If a taxon name belongs to an inclusive taxonomic group, DateLife has two alternative behaviors defined by the "get species from taxon" flag. If the flag is active, DateLife retrieves all species names within the taxonomic group from the standard taxonomy of choice, and adds them to the search query. In this case, subspecific variants are excluded.
If the flag is inactive, DateLife excludes any taxon names above the species level from the search query.
Species and subspecific variant names are processed and searched as provided by the user.
The processed taxon names are saved as an R object of a newly defined class, \texttt{datelifeQuery}, that is used in the following steps. This object contains the input names standardized to a taxonomy of choice (OTT by default), the corresponding OTT id numbers, and the topology of an input tree, if one was provided.

@@

A unique feature of the Phylesystem is that any user can add new published, state-of-the-art chronograms any time, through their curator application (https://tree.opentreeoflife.org/curator). As chronograms are added to Phylesystem, they can be incorporated into the chronogram database of the `datelife` R package, which is currently manually updated as new chronogram data is added to Phylesystem. The updated database is assigned a new version number, followed by a package release on CRAN.
Users can directly run `datelife` functions to trigger an update of their local chronogram database, to incorporate any new chronograms to their DateLife analysis before an official database update is released on CRAN. <!--BCO: What is immediately? Also, it's not just uploading -- they have to include metadata on what brlen mean, maybe? indicate the ingroup/outgroup, do TNRS within OToL, right? Maybe this has improved. I remember being frustrated trying to load in giant trees for datelife use b/c I had to click on the ingroup node.
Luna: Very true, it is not that easy, I added the curation step to hint that there are more steps than simply uploading the chronogram.-->

@@

1. The matching pruned chronograms as newick strings or "phylo" objects.
2. The ages of the root of all source chronograms. These ages can correspond to the age of the most recent common ancestor (mrca) of the user's group of interest if the source chronograms have all taxa belonging to the group. If not, the root corresponds to the mrca of a subgroup withing the group of interest.
3. Study citations where original chronograms were published.
4. A report of input taxon names matches across source chronograms.
5. The source chronogram(s) with the most input taxon names.
6. Various single summary chronograms resulting from summarizing age data, generated using the methodology described next.

@@

**_With no branch lengths.--_**
When producing or obtaining a tree with branch lengths for a group of interest is not possible, DateLife can date a topology without branch lengths, obtained from OpenTree or by implementing the supertree approach described above, by applying the Branch Length Adjuster [BLADJ; @Webb2008; @webb2005phylomatic] algorithm, which requires no initial branch lengths. 
The algorithm starts by fixing ages for nodes with calibration data upon the given topology. Then, it distributes time for nodes with no data evenly between calibrated nodes, minimizing age variance in the resulting chronogram [@Webb2008].
This approach has has proven useful for ecological analyses that require age data [@Webb2008].
When there is conflict between ages of calibrated nodes, BLADJ ignores node ages that are older than the age of a parent node.
The BLADJ algorithm requires a root age estimate to run. If there is no information on the age of the root in the chronogram database, users can provide an estimate from the literature. If none is provided, DateLife will not return a dated topology and will provide a warning message along with suggestions on how the user can provide an age for the root so that a fully dated tree can be returned.
<!-- assigns an arbitrary age to the root as 10% older than the oldest age available within the group. -->

@@

**_Dating a tree with branch lengths.--_**
Relative branch lengths can provide key information for phylogenetic dating, specifically for nodes without any calibration data available. 
While, using initial branch lengths data is the golden standard for phylogenetic dating analyses, producing such trees requires assembling and curating a homology hypothesis, and choosing and implementing a method for phylogenetic inference.
DateLife implements a workflow to streamline this process by applying open data from the Barcode of Life Data System, BOLD [@ratnasingham2007bold] to obtain genetic markers for input taxa.
<!--DNA sequences obtained from BOLD are short (between 500-800 bp) and display little variation. -->
By default, BOLD genetic sequences are aligned with MUSCLE [@edgar2004muscle] using functions from the msa R package [@bodenhofer2015msa]. Alternatively, sequences can be aligned with MAFFT [@katoh2009multiple], using functions from the ape R package [@paradis2004].
The BOLD sequence alignment is then used to obtain initial branch lengths with the accelerated transformation (ACCTRAN) parsimony algorithm, which resolves ambiguous character optimization by assigning changes along branches of the tree as close to the root as possible [@agnarsson2008acctran], resulting in older nodes [@forest2005teasing]. The parsimony branch lengths are then optimized using Maximum Likelihood, given the alignment, the topology and a simple Jukes-Cantor model, producing a BOLD tree with branch lengths proportional to expected number of substitutions per site. Both parsimony and ML optimizations are done with functions from the `phangorn` package [@schliep2011phangorn]. 
Due to the computing load it requires, the BOLD workflow is currently only supported through DateLife's R package. It is not yet available through the web application.

@@

**_Searching the database.--_**
DateLife used the processed input names to search the local chronogram database and found 9 matching chronograms in 6 different studies (Fig. \ref{fig:figure2}B). Three studies matched five input names [@barker2015new; @Hedges2015; @Jetz2012], one study matched four input names [@hooper2017chromosomal] and two studies matched two input names [@barker2013going; @burns2014phylogenetics]. No studies matched all input names. Together, source chronograms provide 28 unique age data points, covering all nodes on our chosen tree topology to date (Table \ref{tbl:table1}).

@@

While we envision that DateLife will have many interesting applications in research and beyond, we recognise that as 
<!-- sort of a "data mining tool",  -->
any other automated tool for biological data analysis, DateLife's results can only be as good as the input data, and as give as good as how we decide to apply them for research.  
<!-- that they might not be appropriate for research. -->
We recognise that this might be a problem in some research areas, but it might not be the case for all. For example, in ecological and conservation biology studies, it has been seen that incorporating some age data provides better results than when not using any age data at all, even if the age data is not accurate [@Webb2008]. In the following sections we discuss the particularities of divergence times resulting from a DateLife analysis for consideration of the readers and users in different research areas.
<!--
Resources of the paragraph above:
https://asistdl.onlinelibrary.wiley.com/doi/full/10.1002/asi.23538?casa_token=3_2mfbffsrsAAAAA%3A42WBe0ffa_fwcff15n4dlwzer5s6nkPCfW1RT8zOmyg0VW9-v_SB5lR8O4uur_MZvl5bUJ0c9FS4qPPTGg
https://towardsdatascience.com/a-primer-on-the-sources-of-biases-in-data-mining-for-machine-learning-d82e89604693

@@

While most chronograms in DateLife's database are constructed using primary calibrations (molecular substitution rates or ages obtained from the fossil record or geological events), DateLife summarizes chronograms using secondary calibrations (ages coming from other chronograms).
@graur2004reading cautioned on the increased error and uncertainty in estimated ages when using secondary calibrations in dating analyses.
@schenk2016sec showed that, in simulations, divergence times inferred using secondary calibrations are significantly younger than those inferred with primary calibrations, when obtained with Bayesian inference methods, and when priors are implemented in similar ways in both analyses.
Accordingly, the scientific community seems to have more confidence in chronograms obtained from a single analysis, using fossil data as primary sources of calibrations [@schenk2016sec], and using fossils that have been widely discussed and curated as calibrations to date other trees, making sure that all data reflect a coherent evolutionary history [@sauquet2013practical], as for example done by @antonelli2017supersmart. <!-- Will this be implemented in future `datelife` versions? Probably not. -->

The Fossil Calibration Database provides availability of 220 primary calibration points encompassing flowering plants and metazoans, that have been curated by experts and used in peer-reviewd publications [@ksepka2015fossil]. The database facilitates the use of expert primary fossil calibrations in new phylogenetic dating analysis. <!--though these still have room to grow... BCO: Polite way to say 'no new additions in four years?' LLSR: lol, yes-->
Despite the wealth of information displayed in the database, users still need the expertise to locate the equivalent node in their phylogenies to be dated. 
Open databases for reuse of fossil calibration data still have room to grow and should be a great addition to the phylogenetics tool set.

@@

Analysis of species diversification using simulated and empirical phylogenies suggest that using a more completely sampled phylogeny provides estimates that are closer to the true diversification history than when analysing incompletely sampled phylogenies [@cusimano2012new; @chang2020estimating; @sun2020estimating]. 
Ideally, phylogenies should be completed using genetic data, but this is a difficult task to achieve for many biological groups.
Hence, DateLife's workflow features different ways of assigning node ages in the absence of calibrations and branch length information for certain taxa.
Completing a phylogeny using a stochastic birth-death polytomy resolver and a backbone taxonomy is a common practice in scientific publications: @Jetz2012, created a chronogram of all 9, 993 bird species, where 67% had molecular data and the rest was simulated; @rabosky2018inverse created a chronogram of 31, 536 ray-finned fishes, of which only 37% had molecular data; @smith2018constructing constructed a chronogram of 353, 185 seed plants where only 23% had molecular data. These stochastically resolved chronograms provide less biased diversification rates estimated with methods that account for sampling fractions [@chang2020estimating; @cusimano2012new], but can also introduce spurious patterns of early bursts of diversification [@cusimano2010slowdowns; @sun2020estimating].

Taxonomy based polytomy resolvers also introduce topological differences.
The study of macroevolutionary processes largely depends on an understanding of the timing of species diversification events, and different phylogenetic and chronogram hypothesis can provide very different overviews of the macroevolutionary history of a biological group. 
For example, alternative topologies in chronograms from the same biological group can infer very different species diversification patterns [@rabosky2015no; @title2016macrophylogenies].
<!-- There is evidence that different chronograms will result in very different macroevolutionary histories inferred from them [@title2016macrophylogenies]. However, it is unknown how using very different age data could affect hypothesis testing in other areas of research and more effort is needed in that direction -->
Similarly there are worries that patterns of morphological evolution cannot be accurately inferred with these types of phylogenies, as any patterns would be erased by randomization [@rabosky2015no]. We note that the same applies for geographical and morphological dependent diversification analysis. Hence, we suggest that these types of randomly resolved phylogenies can be used as null or neutral models, representing the case of diversification independent of traits and geographical scenario.

While stochastic polytomy resolvers have been useful to advance research, notably, risks come with this practice.
Taken to the extreme, one could generate a fully resolved, calibrated tree of all modern and extinct taxa using a single taxonomy, a single calibration, and assigning branch lengths following a birth-death diversification model. Clearly, this can lead to a misrepresentation of the true evolutionary history.
We urge DateLife users to follow the example of the large tree papers cited above, by carefully considering the statistical assumptions being made, potential biases, and assessing the consistency of the results with prior work.

>>>>

authornote: |
  School of Natural Sciences, University of California, Merced, Science and Engineering Building 1, Merced, CA 95340, USA.

@@

Besides binomial species names, DateLife accepts scientific names from any inclusive taxonomic group (e.g., genus, family, tribe), as well as subspecific taxonomic variants (e.g., subspecies, variants, strains). If a taxon name belongs to an inclusive taxonomic group, DateLife has two alternative behaviors defined by the "get species from taxon" flag. If the flag is active, DateLife retrieves all species names within a taxonomic group provided, from a standard taxonomy of choice, and adds them to the search query. In this case, subspecific variants are excluded.
If the flag is inactive, DateLife excludes inclusive taxon names from the search query, and species and subspecific variant names are processed as provided by the user.
The processed taxon names are saved as an R object of a newly defined class, \texttt{datelifeQuery}, that is used in the following steps. This object contains the input names standardized to a taxonomy of choice (OTT by default), the corresponding OTT id numbers, and the topology of an input tree, if one was provided.

@@

A unique feature of the Phylesystem is that any user can add new published, state-of-the-art chronograms any time, through OpenTree's curator application (https://tree.opentreeoflife.org/curator). As chronograms are added to Phylesystem, they can be incorporated into the chronogram database of the `datelife` R package, which is currently manually updated as new chronogram data is added to Phylesystem. The updated database is assigned a new version number, followed by a package release on CRAN.
Users can directly run `datelife` functions to trigger an update of their local chronogram database, to incorporate any new chronograms to their DateLife analysis before an official database update is released on CRAN. <!--BCO: What is immediately? Also, it's not just uploading -- they have to include metadata on what brlen mean, maybe? indicate the ingroup/outgroup, do TNRS within OToL, right? Maybe this has improved. I remember being frustrated trying to load in giant trees for datelife use b/c I had to click on the ingroup node.
Luna: Very true, it is not that easy, I added the curation step to hint that there are more steps than simply uploading the chronogram.-->

@@

1. The matching pruned chronograms as newick strings or "phylo" objects.
2. The ages of the root of all source chronograms. These ages can correspond to the age of the most recent common ancestor (mrca) of the user's group of interest if the source chronograms have all taxa belonging to the group. If not, the root corresponds to the mrca of a subgroup within the group of interest.
3. Study citations where original chronograms were published.
4. A report of input taxon names matches across source chronograms.
5. The source chronogram(s) with the most input taxon names.
6. Various single summary chronograms resulting from summarizing age data, generated using the methodology described next.

@@

**_Dating a tree with no branch lengths.--_**
When producing or obtaining a tree with branch lengths for a group of interest is not possible, DateLife can date a topology without branch lengths, obtained from OpenTree or by implementing the supertree approach described above, by applying the Branch Length Adjuster [BLADJ; @Webb2008; @webb2005phylomatic] algorithm, which requires no initial branch lengths.
The algorithm starts by fixing ages for nodes with calibration data upon the given topology. Then, it distributes time for nodes with no data evenly between calibrated nodes, minimizing age variance in the resulting chronogram.
This approach has proven useful for ecological analyses that require a phylogenetic time context [@Webb2008].
When there is conflict between ages of calibrated nodes, BLADJ ignores node ages that are older than the age of a parent node.
The BLADJ algorithm requires a root age estimate to run. If there is no information on the age of the root in the chronogram database, users can provide an estimate from the literature. If none is provided, DateLife will not return a dated topology and will provide a warning message along with suggestions on how the user can provide an age for the root so that a fully dated tree can be returned.
<!-- assigns an arbitrary age to the root as 10% older than the oldest age available within the group. -->

@@

**_Dating a tree with branch lengths.--_**
Relative branch lengths can provide key information for phylogenetic dating, specifically for nodes without any calibration data available.
While using initial branch length data is the golden standard for phylogenetic dating analyses, estimating trees with branch lengths proportional to substitution rates per site requires obtaining primary data, assembling and curating a homology hypothesis, and choosing and implementing a method for phylogenetic inference.
DateLife implements a workflow to streamline this process by applying open data from the Barcode of Life Data System, BOLD [@ratnasingham2007bold] to obtain genetic markers for input taxa.
<!--DNA sequences obtained from BOLD are short (between 500-800 bp) and display little variation. -->
By default, BOLD genetic sequences are aligned with MUSCLE [@edgar2004muscle] using functions from the msa R package [@bodenhofer2015msa]. Alternatively, sequences can be aligned with MAFFT [@katoh2009multiple], using functions from the ape R package [@paradis2004].
The BOLD sequence alignment is then used to obtain initial branch lengths with the accelerated transformation (ACCTRAN) parsimony algorithm, which resolves ambiguous character optimization by assigning changes along branches of the tree as close to the root as possible [@agnarsson2008acctran], resulting in older internal nodes as compared to other parsimony algorithms [@forest2005teasing]. The parsimony branch lengths are then optimized using Maximum Likelihood, given the alignment, the topology and a simple Jukes-Cantor model, producing a BOLD tree with branch lengths proportional to expected number of substitutions per site. Both parsimony and ML optimizations are done with functions from the `phangorn` package [@schliep2011phangorn].
Due to the computing load it requires, the BOLD workflow is currently only supported through DateLife's R package. It is not yet available through the web application.

@@

**_Searching the database.--_**
DateLife used the processed input names to search the local chronogram database and found 9 matching chronograms from 6 different studies (Fig. \ref{fig:figure2}B). Three studies matched five input names [@barker2015new; @Hedges2015; @Jetz2012], one study matched four input names [@hooper2017chromosomal] and two studies matched two input names [@barker2013going; @burns2014phylogenetics]. No studies matched all input names. Together, source chronograms provide 28 unique age data points, covering all nodes on our chosen tree topology to date (Table \ref{tbl:table1}).

@@

As we envision that DateLife will have many interesting applications in research and beyond, we emphasize that DateLife's results --as well as any insights gleaned from them, largely depend on the quality of the source chronograms: low quality chronograms will produce low quality results. This "garbage in, garbage out" problem has long been recognised in supertree methods for summarizing phylogenetic trees [@bininda2004garbage].
We note that this is a surfacing issue of any automated tool for biological data analysis. For example, DNA riddled with sequencing errors will produce generally poor alignments that will return biased evolutionary hypothesis, independently of the quality of the analysis software used. Again, we urge readers and DateLife users to explore all input chronograms before using a summary chronogram resulting from a DateLife workflow.
Finally, uncertainty and variability of chronogram node age estimates might pose larger issues in some research areas than others. For example, in ecological and conservation biology studies, it has been shown that incorporating some chronogram data provides better results than when not using any age data at all, even if the node ages are not good quality [@Webb2008]. In the following sections we discuss the particularities of divergence times from DateLife's summary chronograms and their impact on certain evolutionary analyses, for consideration of the readers and users in different research areas.
<!--
Resources of the paragraph above:
https://asistdl.onlinelibrary.wiley.com/doi/full/10.1002/asi.23538?casa_token=3_2mfbffsrsAAAAA%3A42WBe0ffa_fwcff15n4dlwzer5s6nkPCfW1RT8zOmyg0VW9-v_SB5lR8O4uur_MZvl5bUJ0c9FS4qPPTGg
https://towardsdatascience.com/a-primer-on-the-sources-of-biases-in-data-mining-for-machine-learning-d82e89604693

@@

While most chronograms in DateLife's database are constructed using primary calibrations (molecular substitution rates or ages obtained from the fossil record or geological events), DateLife also summarizes chronograms using secondary calibrations (ages coming from other chronograms).
The scientific community has generally more confidence in chronograms generated from a single analysis, where fossil data constitute the primary source of calibrations [@sauquet2013practical; @schenk2016sec; @antonelli2017supersmart].
@graur2004reading cautioned on the increased error and uncertainty in estimated node ages when using secondary calibrations in dating analyses, and more recently, @schenk2016sec found that divergence times obtained using secondary calibrations with Bayesian inference methods are significantly younger than those obtained using primary calibrations. While this trend still needs to be evaluated with alternative methods of divergence time estimation,
 many efforts and discussions on best practices for implementing primary calibrations have been advanced in recent years,
for example, using fossils that have been widely discussed and curated as calibrations to date other trees and making sure that all data reflect a coherent evolutionary history [@sauquet2013practical], as for example done by @antonelli2017supersmart. <!-- Will this be implemented in future `datelife` versions? Probably not. -->

Obtaining vetted primary calibrations from the litterature is another bottleneck for facilitating obtention of chronograms.
The Fossil Calibration Database provides availability of 220 primary calibration points encompassing flowering plants and metazoans, that have been curated by experts and used in peer-reviewd publications [@ksepka2015fossil]. This database facilitates the use of expert primary fossil calibrations in new phylogenetic dating analysis. <!--though these still have room to grow... BCO: Polite way to say 'no new additions in four years?' LLSR: lol, yes-->
Despite the wealth of information displayed in the database, users still need the expertise to locate the equivalent node in their phylogenies to be dated.
Open databases for reuse of fossil calibration data still have room to grow and should be a great addition to the phylogenetics tool set.

@@

Analysis of species diversification of simulated and empirical phylogenies suggest that using a more completely sampled phylogeny provides estimates that are closer to the true diversification history than when analysing incompletely sampled phylogenies [@cusimano2012new; @chang2020estimating; @sun2020estimating].
Ideally, phylogenies should be completed using genetic data, but this is a time-consuming and difficult task to achieve for many biological groups.
Hence, DateLife's workflow features different ways of assigning divergence times to taxa with missing the absence of branch length data and calibrations and branch lengths for certain taxa.

Completing a phylogeny using a stochastic birth-death polytomy resolver and a backbone taxonomy is a common practice in scientific publications: @Jetz2012, created a chronogram of all 9, 993 bird species, where 67% had molecular data and the rest was simulated; @rabosky2018inverse created a chronogram of 31, 536 ray-finned fishes, of which only 37% had molecular data; @smith2018constructing constructed a chronogram of 353, 185 seed plants where only 23% had molecular data. These stochastically resolved chronograms provide less biased diversification rates estimated with methods that account for sampling fractions [@chang2020estimating; @cusimano2012new], but can also introduce spurious patterns of early bursts of diversification [@cusimano2010slowdowns; @sun2020estimating].

Taxonomy-based stochastic polytomy resolvers also introduce topological differences in phylogenetic trees.
The study of macroevolutionary processes largely depends on an understanding of the timing of species diversification events, and different phylogenetic and chronogram hypothesis can provide very different overviews of the macroevolutionary history of a biological group.
For example, alternative topologies in chronograms from the same biological group can infer very different species diversification patterns [@rabosky2015no; @title2016macrophylogenies].
<!-- There is evidence that different chronograms will result in very different macroevolutionary histories inferred from them [@title2016macrophylogenies]. However, it is unknown how using very different age data could affect hypothesis testing in other areas of research and more effort is needed in that direction -->
Similarly, there are worries that patterns of morphological evolution cannot be accurately inferred with phylogenies that have been resolved stochastically over a taxonomic backbone, as any patterns would be erased by randomization [@rabosky2015no]. We note that the same applies for geography- and morphology-dependent diversification analysis. Hence, we suggest that phylogenies that have been processed with taxonomy-based stochastic polytomy resolvers, including certain summary chronograms from a DateLife analysis, can be useful as null or neutral models, representing the case of a diversification process that is independent of traits and geographical scenario.

Taxonomy-based stochastic polytomy resolvers have been used to advance research in evolution, still, risks come with this practice.
Taken to the extreme, one could generate a fully resolved, calibrated tree of all modern and extinct taxa using a single taxonomy, a single calibration, and assigning branch lengths following a birth-death diversification model. Clearly, this can lead to a misrepresentation of the true evolutionary history.
We urge DateLife users to follow the example of the large tree papers cited above, by carefully considering the statistical assumptions being made, potential biases, and assessing the consistency of DateLife's results with prior work.
